---
title: 'Linking R and GitHub: part 2'
author: "Simon Brewer"
date: "10/16/2020"
output:
  html_document:
    theme: flatly
    df_print: paged
    number_sections: true
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(png)
library(grid)
```

# R and GitHub

In the previous section, we went through the somewhat laborious steps of setting up your computer to talk to GitHub. We will now look at how to integrate this with R/RStudio, how to set up a R project with GitHub, and the use of GitHub Pages. 

# Setup SSH

[NOTE THAT THIS SECTION MAY NO LONGER BE NEEDED IF YOU HAVE USED GITHUB CREDENTIAL MANAGER]

To avoid having to type in your GitHub password every time you make a request (commit, clone, etc), we can set up an SSH key pair. The key pair consists of a public and private key. The public key can be given to any server you want to connect to. When this is paired with a client that has the private key (your computer), the server unlocks without need for a password. 

To do this, we need to 

- Generate the SSH key pair
- Add the private key to your local SSH client
- Add the public key to GitHub

There are a variety of ways of doing this, but the easiest is to rely on built-in tools in RStudio, so start this up now. 

First, it's worth checking to see if there is already an existing key pair. Go to [Tools] > [Global Options] > [Git/SVN]. If you see something like `~/.ssh/id_rsa` in the SSH RSA Key box, you have existing keys that can be used, and you can skip the next section. If you don't see anything, but you know that there should be a key already present, you can check this by looking at the contents of the `.ssh` directory on your computer:

```
ls -al ~/.ssh/
```

If you see a different key pair present, you can use these later on.

If not, you can generate a key pair by clicking on the [Create RSA Key...] button. RStudio will prompt you for a passphrase, but it is probably easiest to skip this for now and add it later. Click [Create] and RStudio will generate an SSH key pair, stored in the files ~/.ssh/id_rsa and ~/.ssh/id_rsa.pub (and the first of these should be shown in the SSH RSA Key box.

Now let's add the private key to the local ssh agent. You will need to do this from a terminal. Again, the easiest way is to open the terminal in RStudio (go to [Tools] > [Terminal] > [New Terminal]) to open a terminal in the console panel. Here, first type the following to check that `ssh-agent` (the local SSH client) is running:

```
eval "$(ssh-agent -s)"
```
And you should see a process ID like this:
```
# Agent pid 95727
```

Now add the key you have created using the following command:
```
ssh-add ~/.ssh/id_rsa
```

Note that if your key pair has a different name, you will need to use that here. 

Finally, we need to add the public key to GitHub. Go to [Tools] > [Global Options] > [Git/SVN]. If your key pair has the usual name, id_rsa.pub and id_rsa, RStudio will see it and offer to “View public key”. Do that and accept the offer to copy to your clipboard. If your key pair is named differently, open it using a text editor and copy the contents to your clipboard.

Now open a browser, go to [GitHub][ghID] and sign in. Click on the profile picture in the top right hand corner and go to [Settings] > [SSH and GPG keys]. Click [New SSH key] and then paste what you just copied into this box. Give it a title (this can be anything you like) and click [Add SSH key]. 

And that's it! If you've made it this far, everything should be in place for you to connect to GitHub directly as you use R. To test the connection, go back to the terminal on your computer and enter the following:

```
ssh -T git@github.com
```

And you should see a confirmation that the RSA host key has been permanently added. 

# RStudio and GitHub

Now, we can now connect up RStudio with GitHub. There are a couple of ways of doing this, but the following method is pretty reliable. 

## Make a new repository on GitHub

We covered this in the previous lab, but let's do this again. Go to [GitHub][ghID], make sure you are logged in and go to [Repositories]. Click the green button to make a new repository. Call this `myrrepo`, and add a short description (something like `testing RStudio setup`). Make sure the check-box to initialize the repository with a README is checked and click [Create repository]. Once created, click the button [Clone or download], and copy the HTTPS URL (if you click on the little clipboard icon this will get copied to your clipboard).

## Make a new RStudio project

To connect this to RStudio, we need to make an RStudio project. This is a little different from the regular R scripts, and acts like a git repository to store a set of files in a common setting. Open RStudio, and make a new project from [File] > [New Project]. Click on [Version Control], then [Git] (Clone a project from a Git repository). 

On the next screen, add the URL you copied from GitHub to the box labeled [Repository URL]. For the next box, accept the given project directory name (this should be the same as the GitHub repository). The final box gives the directory where the repository will be cloned to on your computer. Make sure this is ok, and that the check-box [Start in new session] is checked. Now click [Create project].

RStudio will appear to restart, launching a new local RStudio Project that connects to the repository on GitHub. It will download the README file, and this should appear in the file browser panel. 

If all this has worked, congratulations - you have successfully linked RStudio to your GitHub account, and can now make changes and push these to GitHub. Note that nearly all these instructions only need to be carried out once. The only part of this you will do routinely is the last part: creating new projects for different bits of work you are carrying out. 

## Start making changes to your project

Let's try and make some changes. First, make sure the README.md file is open in the files panel. If not, click on the file browser, which will show you the contents of the project directory, and double click the `README.md` file. Add a small amount of text to this (I’ll leave this up to you). 

If you now look at the RStudio panel with the Environment, History and Connections tabs, you should see a new one labelled [Git]. Click on this and you should see your READ.md file with a small blue ‘M’ next to it. This indicates that this file has been modified and is different from the version downloaded from GitHub. If you're happy with the changes you made, click on the check-box next to the 'M' (under the heading 'Staged') to stage these changes for uploading.

Now click the [Commit] button. A new window will open showing a) all the files; b) the staged files with a check mark; c) a blank panel to add a commit message; d) a display of which lines have been changed in the currently highlit file (green for additions, red for deletions). Check that your file is still listed as 'Staged' and add a commit message; a short description of the changes. Once this is done, click [Commit] to commit your changes. This effectively creates an updated local version of the repository, including the staged changes. However, if you check your GitHub repository, and look at the README file there, you will see that the changes have not yet been uploaded. To do this, we need to push the changes, so click this button – you should see a short message detailing the upload. Go back to your browser and recheck the README file and you should see your changes appear (refresh the browser page if they are not there). 

Now let’s add an R script. In RStudio, click [File] > [New File] > [R Script] to open a blank R script. Add the following R code to this script:

```{r results='hide'}
## OLS example
x = rnorm(100, mean = 20, sd = 5)
y = x + rnorm(100, mean = 0, sd = 2)

fit = lm(y ~ x)

summary(fit)
```
 
This simply generates some correlated random numbers and builds a linear model between them, and prints out the results of that model. Save the script file as ‘lmExample.R’ and then click ‘Run’ from the scripting window. This will run all the code and allow you to check for mistakes. If all is ok, go through the same steps to commit the changes:

-	Click on [Commit] from the ‘Git’ tab
-	Stage the files you want to push
-	Add a commit message
-	Push to GitHub
-	Check that your changes are on your GitHub site

Let’s add a couple more changes. First, we’ll add code to plot the data, and the model fit to the end of the file: 

```{r fig.keep='none'}
plot(x,y)
abline(fit, col=2)
```

Run to check your code (and the beautiful plot you get), and then stage, commit and push your changes. Make any other changes that you would like to, and when you are done, commit these and then close RStudio. 

# Cloning an existing repository

In this section, we'll look at how you can clone the code from someone else's repository. This will make your own copy of the code that you can then go ahead and modify for your own use. Note that this copy is then independent of the source repository, and *any changes you make will only apply to your copy, not to the original source*. This is mainly used if you want to copy some code for your own purposes, and not collaborate on code development. For this section, you will again need to work in pairs or small groups. To start:

- Make a new repository on GitHub. Make sure this has a different name to that of your working partner
- Clone your repository onto your computer by creating a new RStudio project following the instructions above

Now let's add some R code. The following is just some simple code to make a plot from a file, feel free to replace this with your own code if you'd prefer. 

Rather than a simple script, we'll make a function to make the plots. Open a new R script and enter the following code:

```{r}
plot_ts <- function(x, y, ...) {
  fit <- mgcv::gam(y ~ s(x))
  plot(x, y, type = 'l', ...)
  lines(x, fitted(fit), col = 2, lwd = 2)
}
```

This function will plot some time series data and add a smoothing line based on a GAM fit. Save the script as *plot_ts.R* and push it to GitHub with a suitable commit message. 

Next, we'll add some data to the local repository. 

Next, we're going to add a file to the repository. Get the Mauna Loa CO$_2$ csv file from the slack channel/email chain and add it to the project directory on your computer. If you've forgotten where this is, typing the following command will print R’s current directory 

```{r results='hide'}
getwd()
```

Now if you check your Git tab, you should now see this file, with a small 'A' to indicate this is a new file added to the repository. Click on the check box next to the file name to stage it, then click [Commit] and add a short message. This time, we won't push the changes to GitHub just yet, so close the 'Review Changes' window when you are done. If you look at the [Git] tab in RStudio, you should now see a message stating that `Your branch is ahead of origin/main by 1 commit`. You can make several local commits in this way, and then push them all to GitHub on one go. This can be useful if your connection is limited in anyway.

Finally, we'll add a simple R script that reads the Mauna Loa CO$_2$ file and plots it. Open a new script and add the following code (note we need to load both the data and the file containing the plot function.

```{r fig.keep='none'}
# Load function
source("plot_ts.R")
# Load file
co2 <- read.csv("co2_mm_mlo.csv")
# Make plot
plot_ts(co2$decdate, co2$interpolated, main = "Mauna Loa CO2 record")
```

Save and run your script, and if it all works you should see the following, exciting plot. 

```{r echo=FALSE}
plot_ts(co2$decdate, co2$interpolated, main = "Mauna Loa CO2 record")
```

Now commit and push this to GitHub (note that this will push the previous commit as well). Finally, go back to your browser and check that everything has uploaded correctly. 

If you want to see a list of the changes that you have made (i.e. the equivalent to `git log`), simply click the [History] button in the Git tab in RStudio to show a window with the individual commits (top panel) and the diffs (changes; bottom panel):

```{r echo=FALSE}
img <- readPNG("images/git_history.png")
grid.raster(img)
```

# Cloning another repository

Now clone your partner's repository:

- First, close RStudio
- Go to their GitHub page and find the repository that they have just made. Click the green [Code] button and copy the link.  
- Open RStudio and use the file menu to open a new project. 
- Click on [Version Control], then [Git] (Clone a project from a Git repository). 
- Add the URL you copied from GitHub to the box labeled [Repository URL]. 
- Make sure the the directory where the local copy of the repository will be kept is ok and click [Create project].

The project will now be created and the files will be cloned from GitHub to your computer. 

# GitHub pages

If you click on the name of any of the files in your repository, GitHub will show you the contents as long as they are text files. If you add an HTML document to your repo, GitHub will not try to render it, but simply show you the underlying code. If you want to show the rendered HTML, you need to use GitHub pages instead. This requires a little more set up, but is worth knowing as you can generate HTML pages using R Markdown, then push these and make them publicly visible. 

You can do these steps in one of your existing repositories or create a new one. Either way, make sure that you have an R project that is linked to a GitHub repository. First, in the folder containing your local repository add a new sub folder called `docs`. Next, open an R Markdown file in RStudio, selecting HTML output. Add the following code to the header, just underneath the `title`:

```
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
```

Then modify this file to run some R code and show the results. Save this file and then click the [Knit] button to make sure the HTML compiles. Now stage/commit your changes, *making sure that this includes the `docs` folder*. Now push these changes to GitHub.

Go back now to yoru GitHub repository and click [Settings] from the top menu. Scroll down to the section labelled `GitHub Pages`. Under 'Source', set 'Branch' to `main` and the folder to `docs/` (if you don't see this folder, it was not not uploaded in the commit).

```{r echo=FALSE, fig.asp=0.2}
img <- readPNG("images/github_pages.png")
grid.raster(img)
```

Choose a theme for your web page from under 'Theme chooser'. GitHub will now set up your webpage, and the URL will appear at the top of this section (you might need to refresh the Settings page to see this). Click on this now to open the web page. 

Now if you want to make changes to the webpage, simply update the markdown in your local repository, knit it to HTML and stage/commit/push it to GitHub. Note that if you updated the theme, you will need to pull the changes back to your local repository first. 

# Undoing changes

By far the safest way to 'undo' a change in your git repository is by correcting the problem and committing this with a suitable message. However, sometimes you may need to roll back your changes to a previous commit. git has a couple of methods for doing this: `reset` and `revert`. These are command line tools, so if you want to test them, you will need to open a terminal and change directory (`cd`) to your local repository. 

## `reset`

If you have made local commits but not yet pushed these to GitHub (or any other location), then the `reset` function is fairly straightforward. Each commit that you made is accompanied by a short hexadecimal hash string called SHA1. This can be used directly with the `reset` command to move back to ab earlier version of your code. To see the SHA1 numbers, together with the accompanying commit message, simply type:

```
git log --oneline
```

The output will look something like this, with the SHA1 number in the first column, followed by the message:

```
b764644 Third commit
7c709f0 Second commit
9ef9173 First commit
```

Now to revert to the second commit, we use the corresponding number (`7c709f0`) with the reset function as follows. Note that your numbers will be different, and you should probably not run this unless you are happy to rollback. 

```
git reset 7c709f0
```

If you run `git log --oneline` now, the output would look like this:

```
7c709f0 Second commit
9ef9173 First commit
```

You can also use relative changes, e.g. the following will move back two places from the current commit:

```
git reset current~2
```

## `revert`

The `reset` function acts like the undo function in most pieces of software. It effectively removes any changes that you have made. The `revert` function works in a slightly different way by creating a *new* commit that reverses the previous changes. For example, if you had added a line of code in the last commit, then the `revert` function creates a new commit in which this line is removed. To revert to the previous commit, simply type:

```
git revert
```

Note that this is also an option in RStudio's [Git] tab under the [More] drop-down menu. Because this adds a new commit, this will prompt you to enter a message. 

The main difference is that this retains the full set of previous commits *plus* the new one, allowing a more detailed trace of your changes. If you run the `revert` command on this same example, the output from the log will now look like this:

```
11b7712 Revert "Third commit"
b764644 Third commit
7c709f0 Second commit
9ef9173 First commit
```

In general, this is the preferred method for undoing changes, particularly with repositories that are shared with other people.

# Deleting GitHub repositories

We've made quite a few example repositories so far. If you would like to delete any git repository, you will need to both delete the local copy and the one hosted by GitHub. The local copy can be deleted like any normal folder on your computer. The remote copy can be deleted by going to the repository page on GitHub, clicking on [Settings], scrolling all the way down to the Danger Zone[^1], and selecting [Delete this repository].

[^1]: Sterling Archer approves of GitHub

[g4wID]: https://gitforwindows.org
[gitID]: https://git-scm.com/downloads
[hgwrID]: https://happygitwithr.com
[ghID]: https://github.com
