---
title: 'Linking R and GitHub: part 1'
author: "Simon Brewer"
date: "12/10/2020"
output:
  html_document:
    theme: flatly
    df_print: paged
    number_sections: true
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(png)
library(grid)
```

# **git**

**git** is a version control system designed by the developers of the Linux kernel. It is primarily used to manage the development of source code, but can easily be used to manage any set of files. It is free to use and install and is particularly useful when managing files for collaborative work, including:

- version control to avoid overwriting new code with old
- version storage to find older modifications
- tracking for issues and bugs
- commenting system to make it clear(ish) what changes have been made and why

The whole system works through repositories: a set of files dedicated to a given project. This is not restricted to R files, and can be used with most computer languages, documents and files. Repositories can be stored on your own computer (*local*) or hosted on a server (*remote*). In practice, most people use a combination of these, with your files being changed locally, then *committed* to the hosting service. 

The examples used here are based on the information available at [happygitwithr][hgwrID]. This website has much more extensive instructions on the use of git and R (and git on its own), and is an excellent place to follow up. 

# GitHub

GitHub is one of several online hosting sites, that provide cloud-based storage for your git repositories (other options are GitLab and Bitbucket). This (and git) now integrate quite seamlessly with R, and particularly with RStudio, and we'll go through the steps of setting up R and GitHub to play nicely together. 

The overall setup scheme is as follows:

1. Create an account at GitHub
2. Make sure R/RStudio is up-to-date
3. Install git
4. Link git to your account
5. Connect R to git and GitHub

We'll tackle the first 4 of these in this lab, and go through some basics of working with git on the command line.

## Setting up a GitHub account

Go to this site: https://github.com and follow the instructions. Choose your username wisely - remember that this is way that potential employers can see your work!

## Update R and RStudio

Go to 

- http://cran.r-project.org  for R
- https://www.rstudio.com for RStudio

## Install git

git is the underlying software that manages the repositories, keeps files current and copies and sends files between computers, so we need to install this first so that R can use it. 

### Windows

The easiest way to install git is using [Git for windows][g4wID]. This is a self contained installer; download it, run it and accept the default settings (although feel free to switch to a more modern text editor than vim).

### Mac OSX

The easiest way to install git on a Mac is by installing the XCode command line tools. Open a terminal and enter the following command to do so:

```
xcode-select --install
```

This can take a little time, but when finished will provide you with a wide range of useful tools for programming. 

### Linux 

Use your favorite package manager to install git directly. 

Ubuntu or Debian Linux:

```
sudo apt-get install git
```

Fedora or RedHat Linux:

```
sudo yum install git
```

You can also get the latest version of git [here][gitID]. 

## Setting up git

Once the installation is complete, open a terminal to check that git is working. There are several options to access a terminal, but an easy one is to use the terminal in RStudio. Open RStudio, then go to [Tools] > [Terminal] > [New Terminal], and you should get a new tab in the console panel. Go to this, and type the following to check that git is working:

```
which git
```

and 

```
git --version
```

If this doesn't work, open [Tools] > [Global options]. Select the Terminal from the left hand side, and look at the drop down list titled [Open new terminal with..]. If there is an open for `git bash`, select this and try again. 

On Mac or Linux, you can also use a regular terminal, as long as it is running bash. On Windows, you can access Git Bash from the Start menu, which will open a terminal.

If this works, the next steps are to tell git your name and email address. The email address should be the address that you registered with GitHub, but the name can be your real name. Type the following in your terminal:

```
git config --global user.name 'Grumpy Git'
git config --global user.email 'grumpyoldgit@example.com'
```

To check that this has been correctly entered:

```
git config --global --list
```

# Connect to GitHub

In this section, we will walk through the steps of connecting to GitHub for the first time, and both pulling from and pushing to GitHub from your computer.

## Make a repo on GitHub

Go to https://github.com and make sure you are logged in to your account. Click on [Repositories], then click the green [New] button.

This will open a form with several options. Enter the following:

- Repository name: `myrepo` (or whatever name you will easily remember).
- Description: "`setup test`" (or something else, but it is good to get into practice of writing something descriptive in the README).
- Public. `YES` 
- Initialize this repository with a README. **This is a key option for linking to R**
- For everything else, just accept the default.

Click button [Create repository.]

This has created the remote copy of your repository. We now need to link this to a local copy on your computer. Click the green button [Code], and copy the HTTPS URL (if you click on the little clipboard icon this will get copied to your clipboard).

Go back to the terminal on your computer. Change your working directory until you are somewhere where you are happy to be making a copy of your repository. In a terminal:

- `pwd` prints the current directory
- `cd` changes the directory
    - `cd` will go to your home directory
    - `cd ..` will go up one level
    - `cd /path/to/my/files` will change to whatever path you have listed
- `ls` will print a list of the files in the current directory

If you are working with Windows, the easiest way to create a folder that you can easily find is to change to the desktop:

```
cd /c/User/USERNAME/Desktop
```

And then make a new folder here that you will use for this lab

```
mkdir rgithub
```

Finally change to this directory

```
cd rgithub
```

Once you have changed directory to where you want the local files saved, make a clone of your repository by typing the following into the terminal:

```
git clone https://github.com/USERNAME/REPOSITORY.git
```

Replace the `https://` address with the URL copied from GitHub. git will now make a new directory with the name of your repository and copy all files from GitHub to this directory. You should see something similar to the following output if this has been successful:

```
Cloning into 'myrepo'...
remote: Enumerating objects: 3, done.
remote: Counting objects: 100% (3/3), done.
remote: Total 3 (delta 0), reused 0 (delta 0), pack-reused 0
Receiving objects: 100% (3/3), done.
```

To check that this has worked, we will change directory to the newly created clone repository. and take a look at its contents. 

```
cd myrepo
ls
head README.md
git remote show origin
```

The `head` command will show you the first few lines of the README file. This is a markdown file and is used as the default landing page for your repository. The last of these commands will provide information about where this would be cloned from, information about where the master repository is held, and whether or not the local clone copy is up-to-date (it is for now). 

```
git remote show origin
```

## Make a local change, commit, and push

We will make a simple change to the `READ.md` file, and then check to see if we can push this to GitHub. You can open this file with a text editor, or run the following command line to insert some new text. 

```
echo "This is my exciting first GitHub repo. Whoo." >> README.md
```

If you want to check that git has seen this change, type the following at the command line

```
git status
```

And this should tell you that one file (`README.md`) has been modified. 

```
    modified:   README.md
```

Now let's push this change to git. This is done in three parts. 

```{r echo=FALSE, fig.asp=0.2}
img <- readPNG("images/git_flow.png")
grid.raster(img)
```

Fig: git work flow (https://thenewstack.io/git-with-the-program-getting-started-with-github/)

First we *stage* the changes (basically creating a list of all the modifications we wish to push. Then we create a *commit*, the package of changes along with a brief description of what the modifications are. Finally, we *push* this to git. You should at this point be asked to enter your username and password for GitHub, so do so, and the changes will be pushed to GitHub. 

```
git add -A
git commit -m "A commit from my local computer"
git push
```

The description of changes is a key part of the way git works, and you will not be allowed to push changes without this. The goal here is to make the description brief but informative, and in turn the commits you make should be relatively small increments each time. Remember that the goal of these comments is to allow collaborators and other users to understand what is going on and how this is affecting the code. 

Depending on your operating system, this might open a browser page for GitHub's credential manager, which can be used to store your login details. Here, you need to click the green button [Authorize GitCredentialManager], then enter your password. This may not work in certain browsers (e.g. Windows Edge). If the button is grayed out and you can't click it, close the browser and stop the `push` command (Ctrl-C). Now set you default browser to something else (e.g. Chrome) and try and run the `git push` command again. 

If this has worked, you should see some output similar to this:

```
Enumerating objects: 5, done.
Counting objects: 100% (5/5), done.
Delta compression using up to 8 threads.
Compressing objects: 100% (2/2), done.
Writing objects: 100% (3/3), 311 bytes | 0 bytes/s, done.
Total 3 (delta 0), reused 0 (delta 0), pack-reused 0
To https://github.com/grumpy-old-git/myrepo.git
   b4112c5..de669ba  main -> main
```

## Confirm the commit

Now let's check to see if the local change propagated to the GitHub repository. Go back to your GitHub page and refresh the repository page. You should now see the new line in the README file.

While the README file is automatically shown when you access the directory, you can also look at the content of any other file you have added to the repository. Click on [Code], at the top of the page, then select any file (there will probably only be the README file for now). If the file is a text file, the content will be displayed in the window. This window allows you to track past changes in the file by clicking on the 'History' link on the right hand side. For now, this will just show the initial commit when the repository was made, and the update to the README file. At the far right hand side of this page is a small icon that looks like this `<>` for each commit. Clicking this will show you the files in the repository *before* that commit was made, which allows you to track the changes made by yourself (and others). 

# Pulling changes from GitHub

If there have been any changes to the remote repository, we need to update the local one to reflect these. To do this, we need to git's `pull` function. To demonstrate how this works, first go to your remote repository on GitHub. Scroll down to where the contents of the README.md file are shown, and click on the pencil icon in the top right to edit this. 

```{r echo=FALSE, fig.asp=0.15}
img <- readPNG("images/edit_readme.png")
grid.raster(img)
```

Enter some text and then scroll down and click [Commit changes]. Go back to the main repository page and check that the new text is visible. No go back to your terminal and type 

```
git pull
```

And your changes will be downloaded from the server. In the output from the command thre should be a brief description of the changes:

```
    Fast-forward
     README.md | 2 +-
     1 file changed, 1 insertion(+), 1 deletion(-)
```


Note that if there are no changes, running `git pull` simply stop with the message `Already up to date`, and not modify anything. As a result, there's little risk in running this command, and it's good practice to pull every time that you start working on a repository, particularly if these are collaborative. 

> Whether you haven’t opened RStudio in a month or you’ve just been away for a lunch break, pull. It might not be necessary, but it can save a lot of heartache later.

# Adding files

Next, we'll add a file to the local repository and push this to GitHub. The example here uses a Netlogo model (`randomWalk4.nlogo`), but you can upload any file. Note that GitHub limits the size of the files you can upload to 100Mb, so it's best to choose a small file for this. Copy the file to your local repository (e.g. `Desktop/rgithub/myrepo`). Once it is there, go back to the terminal and enter `git status` to check that git has seen the new file. It should tell you that you have an *untracked* file in the repository. This simply means that git has found the file, but is not tracking it as part of the repository. To fix this, we need to add the file:

```
git add randomWalk4.nlogo
git status
```

The file will now be marked as a *new* file, and will be added the next time you make a commit. So let's do that now. 

```
git commit -m "Added model code"
git push
```

If you want to add several files, then running `git add .` will add everything in the directory that is currently untracked, or you can use wildcards `git add '*.R'`. There may be files that you do not want to commit to GitHub, either because these are too large or because these include sensitive information. You can exclude these by adding them to a file called `.gitignore`. First check if this already exists in your repository

```
ls -a
```

If it doesn't, you can create it using the Unix `touch` command:

```
touch .gitignore
```

As an example, we'll now make a copy of the file that you just pushed to your repository (the new name doesn't matter)

```
cp randomWalk4.nlogo allMySecrets.txt
git status
```

Which lists the new file. Add this to the .gitignore file

```
echo allMySecrets.txt >> .gitignore
git status
```

And this has now been dropped from the list of untracked files. If you want to see your history of commits, simply type `git log` in the terminal to print out the author, date and comment for each one.

# Cloning a repository

The last thing we'll do in this lab is to clone someone else's repository. For this part of the lab, you'll need to work in pairs or a small group. Using the instructions from this lab, each person should

- Make a new repository on GitHub. Make sure this has a different name to that of your working partner
- Clone your repository onto your computer
- Update your README with a *polite* message for your working partner
- Add a file to your local repository and use the add/commit process to push it to GitHub

Once you have done this, clone your partner's repository onto your computer. To do this:

- Go your partner's GitHub page and find the correct repository. 
- Click on the green [Code] button and copy the link
- Go to your terminal and type `cd ..` to change to the higher level directory 
- Clone the repo: `git clone https://github.com/NAME/REPOSITORY.git` (but replace the https link with the link that you copied)

If you have got to this point, you now have a successfully working git installation. We will next look at how to integrate this properly with R/RStudio.

[g4wID]: https://gitforwindows.org
[gitID]: https://git-scm.com/downloads
[hgwrID]: https://happygitwithr.com
